{\bf Log Replay for Code Generation in OLTP Systems}

### 2021-07-22 old1.tex
Updates from Overleaf

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Filter Representation in Vectorized Query Execution}}
\author{Amadou Latyr Ngom}
\date{June}
\Year{2020}
\trnumber{}

\committee{
Andy Pavlo, Chair \\
Todd C. Mowry \\
}

@@

\keywords{Vectorization, Compilation, Filter, SIMD}

@@

\begin{abstract}
In-memory Database Management Systems (DBMSs) use vectorization to reduce interpretation overhead, improve cache locality, and perform loop optimizations. For each \textit{vector} (i.e. batch of tuples), the DBMS needs to track the set of valid tuples after a predicate application. Systems employ one of two representations: Selection Vectors (SelVecs) or Bitmaps. In this work, we analyze each approach's strengths and weaknesses to provide recommendations on how to implement vectorized operations. We determine, through a wide range of micro-benchmarks, that the optimal implementation strategy is a function of many factors: the cost of the iterating through tuples, the cost of the operation itself and how amenable it is to SIMD vectorization. Our analysis shows that Bitmaps perform better for operations that can be vectorized using SIMD instructions but that SelVecs otherwise perform better due to a cheaper iteration logic.
\end{abstract}

\begin{acknowledgments}
My advisor is cool.
\end{acknowledgments}

@@

\chapter{Introduction and Background}
DBMSs typically handle two broad classes of applications: Online Transaction Processing (OLTP) and Online Analytical Processing (OLAP) applications \cite{abadi_phd}. OLTP applications tend to be short-running, write-heavy queries that only touch a few entries in a database. OLAP applications, which are the focus of this thesis, are long-running, read-mostly queries that analyze large amounts of data. Historically, this data was disk-resident; disk accesses were the bottleneck of OLAP queries. As DRAM sizes increase, and the cost per GB decreases, large databases begin to fit in main memory. Rather than optimize for disk access, modern DBMSs must optimize for CPU efficiency by reducing instruction count and cycles per instruction when executing queries.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/SampleQuery.png}
    \caption{\textbf{Sample SQL Query and Plan.}}
    \label{fig:intro_query}
\end{figure}

The traditional execution technique used in most DBMSs \cite{mysql, posrgres}, the Iterator Model \cite{volcano}, became ill-suited for modern OLAP applications. It executes a query by interpreting the query plan. Each database operator implements a virtual \texttt{next()} function that returns the next tuple it can generate; each expression (i.e. addition, multiplication) also provides an virtual \texttt{evaluate()} function that takes in a tuple and returns a value. \cref{fig:volcano_graph} showcases the Iterator model with the sample query in \cref{fig:intro_query}. The operators \texttt{Scan A} and \texttt{Scan B} iterate through the table and emit the tuples to the parent operator. The filter only returns tuples that satisfy the \texttt{WHERE} clause. The join first builds a hash table, then probes it to find matches.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Volcano.png}
    \caption{\textbf{The Iterator Model on the Sample SQL Query.}}
    \label{fig:volcano_graph}
\end{figure}

We can see why the Volcano model is ill-suited for in-memory DBMSs. For every tuple -- of which there can be billions in an OLAP setting -- the virtual function calls (marked in red in the figure) constitute a significant overhead.

To minimize this overhead for in-memory DBMSs, researchers have developed the Vectorization Model.

\iffalse
\section{Compilation}
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/Pipeline.png}
 \caption{Pipelines within the sample query.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/PipelineCode.png}
 \caption{Code generated for the sample query.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{The compilation model with the sample query. Each pipeline and its corresponding code have the same color.}
\label{fig:update_map_intro}
\end{figure}

Query compilation \cite{system_r, hique, hyper_llvm} aims to eliminate the cost of interpretation by compiling the query plan down to machine code.
{
\color{red}
Describe what compilation means.
}

The data-centric compilation technique pioneered by Hyper \cite{hyper_llvm} splits the plan into several pipelines and compiles them individually. A pipeline is a sequence of relational operators that do not materialize the input result set followed by a pipeline-breaker -- an operator that does materialize the input set (e.g., to build a hash table, or to sort tuples). \cref{fig:pipeline_graph} showcases the pipelines of the sample query. The first pipeline-breaker occurs in the join operator because it needs all input tuples to build a hash table. \cref{fig:pipeline_code} lists the code generated for the query. Conceptually, it performs the same task as \cref{fig:volcano_graph} without any of the virtual function calls. In addition to eliminating interpretation, this approach maximizes register/cache efficiency because a pipeline fully processes a single tuple before fetching the next one. Compilation thus reduces both the number of instructions executed and the number cycles per instructions.
\fi

\section{Vectorization}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Vectorization.png}
    \caption{\textbf{The Vectorization Model with the Sample Query} -- The code is nearly identical to that in \cref{fig:volcano_graph}}
    \label{fig:vectorization_graph}
\end{figure}

The vectorization model, pioneered by Vectorwise \cite{vectorwise}, is almost identical to the Volcano model, except that it deals with batches of tuples or \textit{vectors} (e.g., 2048 tuples at a time). The \texttt{next()} function returns a vector of tuples, and the \texttt{evaluate()} function returns a vector of values. \cref{fig:vectorization_graph} showcases vectorization model on the sample query in \cref{fig:intro_query}; it is near identical to \cref{fig:volcano_graph}.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/VectorMul.png}
    \caption{\textbf{A Vectorized Multiplication Function} -- The code in green indicates compiler or programmer optimizations to exploit Intel's AVX512 features.}
    \label{fig:vector_mul}
\end{figure}
Vectorization provides two main benefits. First, it amortizes the cost of interpretation by the size of vectors. Second, most functions that operate on vectors are highly amenable to loop optimizations (e.g., loop unrolling, SIMD vectorization) by the compiler or by the programmer. Consider, for a example, the function shown in \cref{fig:vector_mul}. The iterations of the loop are independent, and the floating-point multiplication possesses a SIMD variant, which means the loop can be unrolled and vectorized by the compiler.

\iffalse
\begin{figure}[H]
    \centering
    \includegraphics[scale=0.5]{images/VectorCompilation.png}
    \caption{A hybrid vectorization and compilation model with the sample query..}
    \label{fig:mix_compilation_vectorization}
\end{figure}
Our system \cite{rof, noise_page} combines both data-centric code generation and vectorization : it generates code that operates on vectors. This approach eliminates interpretation overhead and still enables loop optimizations. \cref{fig:mix_compilation_vectorization} showcases the mixed model. It differs from \cref{fig:compilation_graph} only in that it operates on vectors.
\fi

Vectorization introduces a new challenge. Individual tuples within a vector may be \textit{filtered out}, meaning that should not be processed. Only \textit{selected} tuples -- those not filtered out -- should be processed. For example, in \cref{fig:vectorization_graph}, only tuples that satisfy the \texttt{WHERE} clause should be inserted in the hash table. Next, we discuss the methods used to represent filters, i.e., to identify selected tuples within a vector.

\section{Filter Representation}
\cref{fig:repr_intro} displays the two common ways to represent filters: Selection Vectors (SelVecs) and Bitmaps. SelVecs index into the selected tuples; tuples without an index are implicitly filtered out. For Bitmaps, on the other hand, the set bits indicate the selected tuples; the unset bits indicated filtered out ones.
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.7]{images/FilterRepresentationIntro.png}
    \caption{\textbf{Filter Representation} -- Selection Vectors (left) index into the tuple vector. Bitmaps (right) are only set on selected rows.}
    \label{fig:repr_intro}
\end{figure}
Previous work has assumed one of these two representations, without providing a reason for that choice. Vectorwise \cite{vectorwise}, and work derived from it \cite{miro_adapt, everything_vectorized, sompolski_vec},  rely on SelVecs, whereas the more recent VIP \cite{orestis_bitmap} relies on Bitmaps. This paper aims to analyze the trade-offs between these two representations and derive insights for developers implementing vectorized query processing.

\chapter{Computing on Filtered Vectors}
\section{Operations on Filtered Vectors}
\begin{figure}[t!]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/UpdateIntro.png}
 \caption{Update operation.}
  \label{fig:update_intro}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MapIntro.png}
 \caption{Map operation.}
  \label{fig:map_intro}
\end{subfigure}
\caption{\textbf{Operations on Filtered Vectors} -- Updates change the set of selected tuple. Maps apply a function to the set of selected tuples without updating the set.}
\label{fig:update_map_intro}
\end{figure}
\cref{fig:update_map_intro} shows the two kinds of operations on vectors of tuples: Updates and Maps. Updates apply a new predicate function to the vector and update the selected tuples accordingly. A scan filter (i.e., WHERE clause) is an example of an update operation. Maps, on the other hand, keep the selected set but compute a new vector using a mapping function, or reduce the vector to a single value (e.g., for aggregation). A projection (i.e., SELECT clause) is an example of a map operation. All relational operations considered in this paper are decomposable into a set of maps and updates. \cref{fig:probe_example} shows the example of join probes. First, we compute the hashes of the input tuples; the result is a vector of hash values (Map). Then, we perform a hash table lookup using the hashes to obtain a vector of hash table entries (Map). Due to hashing collisions, the entries may not match the input, so we need to compare the join keys to filter out inexact matches (Update). We then advance the hash table chains (Map) and repeat the comparison until entries are exhausted.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/JoinProbeExample.png}
    \caption{\textbf{Multi-step Join Probe} -- Bold squares indicate the hash table entries currently pointed to.}
    \label{fig:probe_example}
\end{figure}

Other relational operators are similarly decomposable. The next section briefly discusses transitioning between steps before we move on to the core of the thesis: optimizing each step.

\section{Transitioning Between Steps}
In this section, we briefly talk about the cost of transitioning between steps. Most of the time, this cost is non-existent because the filter: the transition is \textit{implicit}. The hashing phase and the lookup phase in \cref{fig:probe_example} have the same selection vector. On the other hand, the Chain Following phase operates on entries that do not satisfy the comparison: the transition involves a set difference. Transitions in decomposed disjunctive filters \cite{pcq} (shown in \cref{fig:disjunctive_filter_howto}) involve both a set difference for short-circuiting and a union to get the final result.
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DisjunctionHowto.png}
    \caption{\textbf{Transitioning Between Steps} -- We apply the first clause on all input tuples. In the second step, however, we only apply the second clause on those tuples that don't pass the first clause. In the third step, we take the union between the two outputs.}
    \label{fig:disjunctive_filter_howto}
\end{figure}
To the best of our knowledge, all transitions take the form of set differences (for elements that do not pass the previous predicate) and unions (to obtain the final result). These operations are cheap to implement with Bitmaps because we can easily rely on SIMD instructions. Efficiently implementing set operations on SelVecs is more complicated.  Whenever necessary, Updates can write to one additional SelVec for tuples that do not pass the filter; this second SelVec implicitly corresponds to the result of a set difference. Note that it does not overlap with the first SelVec, so a union of any of their subsets is just an array append, a fast operation. \cref{fig:disjunctive_filter_perf} shows that the overhead of set operations is negligible for both bitmaps and selection vectors. The rest of the paper, therefore, focuses on optimizing the individual steps.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{eval/transition.pdf}
    \caption{\textbf{Transition Overheads} -- The overheads are $3.55\%$ and $4.06\%$ for Bitmaps and SelVecs respectively.}
    \label{fig:disjunctive_filter_perf}
\end{figure}

\section{Compute Strategies}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/ComputeIntro.png}
    \caption{\textbf{Compute Strategies} -- Selective Compute (left) only considers selected tuples. An \texttt{x} indicates an uninitialized value. Full Compute (right) considers all tuples.}
    \label{fig:compute_intro}
\end{figure}
 A compute strategy determines how to perform the Update or Map operations. We consider three compute strategies: Selective, Full, and Mixed Compute. As their names indicate, Selective Compute only applies the given functions to selected tuples, whereas Full Compute applies them to all tuples. \cref{fig:compute_intro} shows each approach's effect. While Full Compute usually performs more work, it greatly benefits from SIMD vectorization, simple loop structure (for loop unrolling and interleaving), and easy branch prediction. To exploit this trade-off, the third strategy, namely Mixed Strategy, switches from Full Compute to Selective Compute when the selectivity (i.e., the ratio of selected tuples) goes below a certain threshold. The next sections will show how to obtain this threshold.

Note that for Update operations, Full Compute is not nicely compatible with SelVecs because a slow intersection of sorted sets \cite{sorted_set} would be required to compute the final selection. This limitation will become important when choosing a strategy to implement. Furthermore, operations with side-effects (e.g., hash table build, aggregation) cannot use Full Compute for correctness reason. This thesis focuses on side-effect free operations.

Furthermore, for SelVecs, there is another dimension we did not consider: Branching Store versus Non-Branching Store. We do not focus on this dimension, for it has been studied extensively before \cite{miro_adapt}. All experiments below use the optimal branching strategy.

\section{The Filter Decision Tree}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DecisionTree.png}
    \caption{\textbf{The Filter Decision Tree.}}
    \label{fig:filter_decision_tree}
\end{figure}
\cref{fig:filter_decision_tree} shows the filter decision tree. For every kind of operation on filtered tuples, the tree indicates what the best representation and compute strategy pair is. It assumes that the developer implements all representations and compute strategies, which is a fairly costly endeavor engineering-wise. When this is not affordable, systems designers can still use this tree as a starting point before making compromises. The remaining part of this section will give a high-level rationale behind the tree. Chapter 4 will provide practical case studies that justify the shape of this tree. Chapter 5 will use the insights provided by this tree to execute an end-to-end multi-step query and show the performance gains.

\section{High Level Rationale}
\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleFull.png}
 \caption{Full Compute.}
  \label{fig:rationale_full}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%  
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleBitmap.png}
 \caption{Selective Compute with a Bitmap. The iteration logic is adapted from \cite{bitmap_iteration}. The SIMD version is equivalent to Full Compute.}
  \label{fig:rationale_bitmap}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%  
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleSelVec.png}
 \caption{Selective Compute with a SelVec. The last \texttt{scatter} is only necessary to maintain consistent indices between the input and output vector during multi-step Map operations. It should otherwise be avoided due to its slowness.}
  \label{fig:rationale_selvec}
\end{subfigure}
\hspace*{\fill}%
\vspace*{8pt}%

\caption{\textbf{Implementing Vector Primitives} -- The code in green indicates the core operation (multiplication). The code in red indicates iteration logic.}
\label{fig:rationale_code}
\end{figure}

Every operation on filtered tuples is a loop, the body of which performs the actual computation. Three factors influence this loop's performance: the iteration logic, number of processed tuples, and time per processed tuple. \cref{fig:rationale_code} shows the three ways to multiply a filtered vector by a constant (a Map operation). Full Compute (e.g., \cref{fig:rationale_full}), regardless of representation, has the cheapest iteration logic because it does not need to differentiate selected indices. For Selective Compute, iterating over a Bitmap (e.g., \cref{fig:rationale_bitmap}) is more expansive than iterating over a SelVec (e.g., \cref{fig:rationale_selvec}) because of the bit operations required. On the other hand, Full Compute processes more tuples than Selective Compute, especially when the selectivity is low. Finally, SIMD instructions usually reduce the number of iterations and the number of cycles per processed tuples. SIMD code is cheapest with Full Compute; Selective Compute requires \texttt{gather} and possibly \texttt{scatter} to differentiate selected tuples.

We can use these these three factors to explain the shape of the decision tree.

\textbf{The root node}: The root of the tree addresses the case of operations on variable-length data, as is often the case for string operations (e.g., string comparison, sub-string). The number of cycles dedicated to each tuple usually far dominates the total cost. In this scenario, Full Compute is highly inefficient because it processes more tuples than Selective Compute. Also, the cost iteration logic is negligible, making representation irrelevant. We, therefore, choose to use Selective Compute, with either SelVecs or Bitmaps.

\textbf{The branching node}: We then move on to the Branching node, which addresses operations with branches in them (e.g., logical operations, \texttt{CASE} statements, \texttt{IN} statements). Branching increases the cost of SIMD relative to SISD: SISD skips unnecessary instructions, SIMD executes all of them using bit-masks. Furthermore, branches within a loop complicate loop optimizations (e.g., unrolling, interleaving). As a result, Selective Compute beats Full Compute. SelVecs only win over Bitmaps because of the simplicity of their iteration logic.

\textbf{The non-simdable node}: We now consider the non-simdable node. The most (only?) relevant operations that qualify are integer division and modulo. Complex pointer manipulations, while technically simdable using expensive scatters and gathers, also fall under this category. No compiler will auto-vectorize them, and few programmers will write and maintain such SIMD code. Like with variable-length instructions, Full Compute would introduce too much extra computation, and the representation does not matter.

\textbf{The SNBS nodes}: The most interesting nodes of this tree are those dedicated to scalar non-branching simdable (SNBS) operations; all three factors simultaneously play an essential role. Besides, for Updates, Full Compute is incompatible with SelVecs; it may be necessary to switch both compute strategy and filter representation depending on the task. The next section will justify the claims in these nodes and provide an efficient way to switch compute strategy and representation.

\chapter{Experimental Evaluation}
In this section, we experimentally justify the shape of the tree. In doing so, we will also provide a general technique that systems can use optimally switch between strategies. \cref{tab:strategies} summarizes all the strategies we implemented. We run all experiments on an Intel Xeon Platinum 8124M CPU @ 3.00GHz with AVX512 support.

\begin{table}[t!]
\centering
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Strategy Name} & \textbf{Representation} & \textbf{Compute} & \textbf{SIMD}      & \textbf{Compatibility} \\ \hline
SelVecPartial         & Selection Vectors       & Selective        & None               & Update and Map         \\ \hline
SelVecManual          & Selection Vectors       & Selective        & Manually Written   & Update and Map         \\ \hline
BitmapPartial         & Bitmaps                 & Selective        & None               & Update and Map         \\ \hline
BitmapFull            & Bitmaps                 & Full             & Auto Vectorization & Update and Map         \\ \hline
BitmapFullManual      & Bitmaps                 & Full             & Manually Written   & Update and Map         \\ \hline
Full                  & Either                  & Full             & Auto Vectorization & Map                    \\ \hline
FullManual            & Either                  & Full             & Manully Written    & Map                    \\ \hline
\end{tabular}
\caption{\textbf{Implemented Strategies} -- For the SIMD columns, `Auto Vectorization' indicates that we rely on compiler auto vectorization, whereas `Manually written' indicates that we write SIMD code ourselves using intrinsics. On the compatibility column, Update operations cannot use both Full Compute and Selection Vectors, which explains the last two entries.}
\label{tab:strategies}
\end{table}

\section{Variable Length Data}
\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_string_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE country < `Japan'}}
  \label{fig:varlen_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_string_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT POSITION(`a' in country)}.}
  \label{fig:varlen_map}
\end{subfigure}
\caption{\textbf{Variable Length Data} -- Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
\label{fig:varlen_map_update}
\end{figure}

In this section, we analyze the performance profile of operations on variable-length data. We choose cheap Update and Map operations -- string comparison, and character location on short strings (country names) -- to show that time spent per tuple typically dominates the cost of string operations. Here, the primitives process one tuple at a time; it is impossible to use SIMD registers on a vector with variable-length data.

The profiles are shown in \cref{fig:varlen_map_update}. The results confirm the high-level intuition described in the previous section. The string operations themselves are so expensive that representation matters little, and full compute is not viable. As a result, we decide to use that Selective Compute with either representation.

\section{Branching Operations}
\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_and_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \&\& col2 < val2}}.
  \label{fig:logical_and_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_and_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \&\& col2 < val2}.}
  \label{fig:logical_and_map}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_or_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 || col2 < val2}}.
  \label{fig:logical_or_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_or_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 || col2 < val2}.}
  \label{fig:logical_or_map}
\end{subfigure}

\caption{\textbf{Branching Operations} -- Each operation contains one branch because of boolean short-circuiting.}
\label{fig:logical_map_update}
\end{figure}

\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_and_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \& col2 < val2}}.
  \label{fig:bitwise_and_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_and_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \& col2 < val2}.}
  \label{fig:bitwise_and_map}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_or_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 | col2 < val2}}.
  \label{fig:bitwise_or_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_or_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 | col2 < val2}.}
  \label{fig:bitwise_or_map}
\end{subfigure}

\caption{\textbf{Bitwise Operations} -- The increase in performance is due to the lack of branching.}
\label{fig:bitwise_map_update}
\end{figure}

We now consider operations involving branching execution paths. We choose the logical or/and operations because boolean short-circuiting involves a single branch to avoid evaluating unnecessary conditions. To isolate the effect of branching, we also evaluate branch-less bitwise or/and operations. We have confirmed that the compiler manages to auto-vectorize the operations.

>>>>

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Log Replay for Code Generation in OLTP Systems}}
\author{Tianlei Pan}
\date{June}
\Year{2021}
\trnumber{}

\committee{
Andy Pavlo, Chair \\
Wenting Ye \\
}

@@

\keywords{Code Generation, Log Replay, Recovery, Query Compilation}

@@

\begin{abstract}
Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records. We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
\end{abstract}

\begin{acknowledgments}
yes
\end{acknowledgments}

@@

\chapter{Introduction}
A \textit{database management system (DBMS)} is a software that is responsible for storing data, analyzing data, and interacting with applications. Depending on the use case, a DBMS can be designed to focus more on either capturing data, or analyzing data. \textit{On-line Analytical Processing (OLAP)} applications focus on reading, analyzing and aggregating cold data that is less likely to be modified. On the other hand, \textit{On-line Transaction Processing (OLTP)} applications work with write-heavy transactions that modify the database frequently.

Consequently, recovery and durability in OLTP applications are more of a pressing issue compared to OLAP systems. Naive or manual backups are usually sufficient for read-only OLAP applications. However, since OLTP databases are being updated by transactions constantly, naive backups will either cause the database to lose significant amount of data, or fail to preserve the \textit{ACID (atomicity, consistency, isolation, and durability)} properties of a DBMS. This requires OLTP applications to develop specialized recovery mechanism to handle these problems. \textit{Write-ahead logging (WAL)} is currently the most widely adopted recovery mechanisms for OLTP applications that preserves ACID properties.

In addition, OLTP systems also require specialized algorithms and implementations to execute write transactions. Query compilation is an important optimization technique used in OLTP systems to greatly reduce the number of instructions that need to be executed. A DBMS uses a processing model (e.g. Iterator, Vectorization, Materialization) to determine how the DBMS executes a query plan. The processing model is responsible for translating a query into an intermediate expression. The execution engine then breaks down the expression into various tasks for execution. To speed up this process, the DBMS can use an optimization technique known as \textit{Query Compilation}. Query compilation requires the DBMS to generate specialized code for execution tasks to increase execution throughput. These specialized code are often created using an off-shelf compiler into a C/C++ program that implements the functionality of a certain query. For instance, HyPer \cite{hyper_llvm} uses the LLVM compiler framework to translate queries into machine code.

While query compilation has proved successful in increasing the efficiency of a DBMS, it comes with several drawbacks. Implementing a query compilation system can be quite an ordeal. It requires additional knowledge of compiler systems (e.g. LLVM) and huge amounts of engineering work to translate different execution tasks. Moreover, the DBMS also need to modify other components to accommodate with an execution engine reliant on query compilation.

One of such components is the recovery system. The structure of log records are vastly different from query plans. Therefore, to implement a recovery system in a query compilation environment, the DBMS would need to implement an additional execution engine to interpret the log records. Instead, we propose a system design that will extend the query execution engine to create custom programs for replaying log records.

\chapter{Background}
\section{Log Replay}
Database recovery systems are responsible for preserving the ACID principles of a DBMS in case of failure. Early recovery systems assume that there is no non-volatile memory. Mention Tandem here ($https://link.springer.com/chapter/10.1007/3-540-51085-0_43$). Modern DBMS can utilize a much larger memory. This allows logging protocols to be simplified (e.g. tracking dirty pages is no longer required). A modern recovery system consists of two parts: logging and recovery.

Logging is the action of storing transactional information on disk. These information are stored in a special data structure called a \textit{log record}. A log record can either contain physical information (e.g. changes made to a specific physical address) or higher-level information (e.g. user-input query). Mention MSSQL (time-travel table).

Recovery is the action of restoring the database system into a consistent state. If available, the DBMS first attempts to restore its previously recorded checkpoint. The database then initiates the log replay process that goes through every log record stored since that checkpoint. SiLO (checkpoint).

Describe how log replay works in Noisepage. (Big Picture)
Logging: Transactions are logged prior to commit.
Recovery: Independent recovery thread that loops over all the log records. The log record is either a redo record or a delete record. The contents of the record are then interpreted into either a staged write or delete. Initialization of catalog is hard-coded.

\section{Query Compilation}
part for Talk about query compilation / code generation. Mention Efficiently Compiling Efficient Query Plans
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/Pipeline.png}
 \caption{Pipelines within the sample query.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/PipelineCode.png}
 \caption{Code generated for the sample query.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{The compilation model with the sample query. Each pipeline and its corresponding code have the same color.}
\label{fig:update_map_intro}
\end{figure}

Query compilation aims to eliminate the cost of interpretation by compiling the query plan down to machine code.

Talk about related work. (Revise this section).
The data-centric compilation technique pioneered by Hyper \cite{hyper_llvm} splits the plan into several pipelines and compiles them individually. A pipeline is a sequence of relational operators that do not materialize the input result set followed by a pipeline-breaker -- an operator that does materialize the input set (e.g., to build a hash table, or to sort tuples). \cref{fig:pipeline_graph} showcases the pipelines of the sample query. The first pipeline-breaker occurs in the join operator because it needs all input tuples to build a hash table. \cref{fig:pipeline_code} lists the code generated for the query. Conceptually, it performs the same task as \cref{fig:volcano_graph} without any of the virtual function calls. In addition to eliminating interpretation, this approach maximizes register/cache efficiency because a pipeline fully processes a single tuple before fetching the next one. Compilation thus reduces both the number of instructions executed and the number cycles per instructions.

Describe how query compilation works in Noisepage. (Big Picture).

\chapter{Method}

Our proposal is to extend the recovery layer and query compilation layer to support the interpretation of log records. This approach can be realized in three steps:
\begin{enumerate}
\item Extract information from log records.
\item Convert the information into a format that can be accepted and compiled by the code generation engine.
\item Execute the compiled code.
\end{enumerate}

\subsection{Log Record}
The log record structure in Noisepage follows closely to the typical format of a write-ahead log. The header of a log record stores metadata about the log record itself (e.g. record type, record size). The header is then followed by a byte buffer that stores the persisted contents. The size of the byte buffer is also aligned to 8 bytes to ensure the entire log record is aligned to 8 bytes.

During a step of the conventional recovery process, the recovery manager only needs to check the type of the log record and proceed with an insert, update or delete operation. For our method, besides the log record type, we need to find out the contents of the log record as well. This requires additional access to the database catalog.

[Should I expand on the specific types here or later???]
\subsection{Compilation Pipeline}
In Noisepage, a query is first converted into a plan node. The execution engine then translates the plan node into Table Producing Language (TPL) code via a translator object. Finally, the TPL code is compiled into native C++ code. [Also need to explain execution context].

The structure of a log record is different from a query in a number of ways. Firstly, the log record only stores physical information of a committed transaction. For example, an insert query requires the execution engine to retrieve a new tuple slot. For an insert log record, the tuple slot is already been determined and stored within the log record.

These differences require subtle modifications to the compilation process.
Describe Plannodes, Translators that are required for Query Compilation. Can use a graph here.

Describe Execution Context, VM, Bytecode Emitter/Handlers. Interpret vs Compiled.

\subsection{Reinterpreting Log Records}
Describe the process of extracting contents from the log records and converting them into plannodes, translators.

For all records: Find out the mapping from physical column ids to logical column ids.

\subsubsection{Insert Redo Record}
Iterate over the columns of a redo record. Find the value types for each column. Get the value using the previously found value types and column id mappings. Transform each value into an expression value type (ConstantValueExpression/ParameterValueExpression). All the values and metadata are then passed into an Insert Plannode, which is then compiled into an executable query. The tuple slot of the insert operation is retrieved from the execution context, which is then passed to the recovery manager.
\subsubsection{Update Redo Record}
TBD. Should be similar to insert/delete.
\subsubsection{Delete Record}
How to delete index? No information is given on the index, except for the tuple slot to be deleted.

Unless we store a Delta in the delete record, there is no way to get the values.

Pass the delete tuple slot into the execution context, which can be processed in the Bytecode Handler to force the delete operation on the designated tuple slot.

\section{Caching Optimization}
Caching Identifier: Database Oid + Table Oid.
\subsection{Metadata}
Metadata can be cached. Acquire metadata from Catalog. Contents need to be invalidated on drop table?
\subsubsection{Physical to Logical Column Mapping}
Required since the recovery protocol is physical, while query compilation uses logical ids.
\subsubsection{Column Value Types}
This is only for Redo Records. This is required to memcpy the correct values out of a log record into a plan node.
\subsection{Compiled Query}
The compiled query can be cached. Mention that there are two execution modes: Interpret vs Compiled.

For Insert Redo Records: Use ParameterValueExpressions for insert plannodes. Then after compilation, swap in the values to insert into the compiled query into the parameter slots.

For Update Redo Records: ???

For Delete Records: Tuple slot is carried by Execution Context, so not much needs to be done here.
\chapter{Experimental Evaluation}
Mention how the benchmark is implemented.
\section{Throughput Measurement}
Use proper names. Machine specifications. FIO benchmarks. Run experiments with different perspectives.

1. Standard Insert Workload. Single database, single table. 100 percent inserts. Max Columns of 5. 5 statement per txn. Transaction length 5. Allows Varlen.

2. HighStress Insert Workload. Single database, single table. 100 percent inserts. 5 statement per txn. Max Column of 1. Does not allow Varlen.

\chapter{Related Work}
1. A Recovery Algorithm for A High-Performance Memory-Resident Database System. An early paper for disk-based DBMS recovery. Proper log replay infrastructure and checkpoint recoveries. https://15721.courses.cs.cmu.edu/spring2018/papers/12-logging/p104-lehman.pdf

2. Different Logging Schemes. Physical Logging vs Logical Logging. Noisepage uses physical logging. Examples: SiloR (https://15721.courses.cs.cmu.edu/spring2020/papers/10-recovery/zheng-osdi14.pdf). It also uses checkpoints, which does not exist in Noisepage.

3. Different Recovery Schemes. Time-travel table (Azure SQL Database https://15721.courses.cs.cmu.edu/spring2020/papers/10-recovery/p2143-antonopoulos.pdf). Os-fork Snapshots (Hyper, https://dl.acm.org/doi/10.1109/ICDE.2011.5767867)

4. Efficiently Compiling Efficient Query Plans
for Modern Hardware. Early paper abouto query compilation using LLVM framework. https://15721.courses.cs.cmu.edu/spring2020/papers/14-compilation/p539-neumann.pdf

\chapter{Conclusion}
